---
title: "Titanic Prediction with Bayes"
author: "Chris Mulvihill"
date: "June 2021"
output: html_notebook
---

# The Approach

In the examples of analysis of this data set that I have seen, there is
a pattern of the output being a logical true/false statement of whether
someone had survived. This generally makes sense since this is using a
well known data set that that the results can be checked against to see
if the prediction was correct. However, this is often not the case. In
my studies, I have found that it is better to include how certain we are
of the results rather than simply including the true/false output.

Using Bayesian statistics within the model, it will be possible to
predict the likelihood of a passenger surviving based on the
probabilities found in the training data.

Using variations of the following formula for each variable in the data
set, it will be possible to compute the base probabilities that can then
be combined for each combination that a passenger fits into.

$$
P\left(survived|female\right)=\frac{P\left(female|survived\right)*P\left(survived\right)}{P\left(female\right)}
$$

For example, if a passenger if female with a first class ticket, we
combine the probabilities for survival given that they are female with
the probability that they have a first class ticket since each factor
may contribute to their survival.

# Initial EDA

Importing the data:

```{r Importing The Data Files, message=FALSE, warning=FALSE}
# loading the tidyverse package
library(tidyverse)

# reading in the training data
titanic.train <- read_csv("../data/train.csv")

# reading in the test data
titanic.test <- read_csv("../data/test.csv")
```

Checking that the data was read in correctly
First 10 rows of titanic.train
```{r first 10 lines of train, echo=TRUE}
head(titanic.train, n=10)
```
First 10 rows of titanic.test
```{r first 10 lines of test, echo=TRUE}
head(titanic.test, n=10)
```

## combining the data sets for cleaning
### Step 1: make the columns match

* Check that the names are the same
* add a "survived" column to the test set

```{r finding different names}
train.names <- names(titanic.train)
test.names <- names(titanic.test)
setdiff(train.names, test.names)
```
```{r finding same names}
intersect(train.names, test.names)
```

The column names are the same with the exception of the "Survived" column that is missing from the test data.

```{r adding survived column to the test data}
titanic.test$Survived <- NA
head(titanic.test)
```

### Step 2: Combine the data sets
using rbind to combine the data into a single set.

```{r combine the sets}
titanic.full <- rbind(titanic.test, titanic.train)
nrow(titanic.full)
```

## Cleaning the Data
### Cleaning "Age"
Looking for missing data in the *Age* column
```{r finding missing age values}
table(is.na(titanic.full$Age))
```
There are 263 missing values in the Age data. 

Writing a function to find the mode
```{r find the mode}
getMode <- function(x){
  keys <- na.omit(unique(x))
  keys[which.max(tabulate(match(x, keys)))]
}
```

Getting the mode of "Age"
```{r finding Age Mode}
age.mode <- getMode(titanic.full$Age)
age.mode
```

Replacing the missing age with the mode for age
```{r replacing missing Age}
titanic.clean <- titanic.full %>% 
  mutate(Age = ifelse(is.na(Age), age.mode, Age))
table(is.na(titanic.clean$Age))
```
All NA values in "Age" have been replaced with the mode of "Age".

### Cleaning The "Fare" data
Finding missing "Fare" values
```{r find missing Fare}
table(is.na(titanic.full$Fare))
```
Only 1 missing Fare

Finding the Mode of the Fare to replace the missing value
```{r Find Fare Mode}
fare.mode <- getMode(titanic.full$Fare)
fare.mode
```

Replacing the missing Fare with the mode
```{r replacing missing fare}
titanic.clean <- titanic.clean %>% 
  mutate(Fare = ifelse(is.na(Fare), fare.mode, Fare))
table(is.na(titanic.clean$Fare))
```

Fare is now clean. 

### Cleaning Pclass data
```{r finding missing data in Pclass}
table(is.na(titanic.full$Pclass))
```
No missing values in Pclass

### Cleaning Sex data
```{r finding missing data in Sex}
table(is.na(titanic.full$Sex))
```
No missing values in Sex

### Cleaning the Embarked data
```{r finding missing values in Embarked}
table(is.na(titanic.full$Embarked))
```
There are 2 missing values in Embarked. 

Finding the most common value to replace the missing data with:
```{r table of Embarked}
table(titanic.full$Embarked)
```
"S" is the most common value so I will replace the missing data with "S". 

```{r replacing missing Embarked data}
titanic.clean <- titanic.clean %>% 
  mutate(Embarked = ifelse(is.na(Embarked), "S", Embarked))
table(titanic.clean$Embarked)
```

The 2 missing values have been replaced with "S".

### Cleaning the Cabin data
```{r viewing the cabin data}
titanic.full %>% 
  filter(!is.na(Cabin)) %>%
  .$Cabin %>%
  head(n=50)
```

The Cabin values all start with a letter which is likely associated with a section on the ship.  The way modern airplane or theater tickets are laid out.  If the letter is extracted from each cabin id and placed in a new "Cabin_Section" vector, it can be used to further classify passengers. *The theory that is that some cabins sections may have been closer to the life boats. 

```{r splitting the cabin letter from number}
titanic.clean <- titanic.clean %>% 
  separate(Cabin, into = c("Cabin_Section", "Cabin_Number"), "\\d", extra = "merge")
```
```{r table of the cabin sections}
table(titanic.clean[!is.na(titanic.clean$Cabin_Section), "Cabin_Section"])
```

```{r missing cabin secion data}
table(is.na(titanic.clean$Cabin_Section))
```
there are too many NAs to fill for this data to be accurate. 


# Determining Initial Probabilites
Probability of Survival from the test data
```{r survived probability}
p_survived <- titanic.clean %>%
  filter(!is.na(Survived)) %>% 
  pull(Survived) %>%
  mean()
p_survived
```


## Probability of survival given sex:
$$P\left(survival\ |\ sex_{i}\right) = \frac{P\left(sex_{i}\ |\ survival\right)* P\left(survival\right)}{P\left(sex_{i}\right)}$$

```{r probability of a passenger being female}
p_female <- mean(titanic.clean$Sex[1:891] == "female")
p_female
```

```{r probability of surviving}
p_survived <- titanic.clean %>%
  filter(!is.na(Survived)) %>%
  pull(Survived) %>%
  mean()
p_survived
```


```{r probability of female given survived}
p.female.survived <- titanic.clean %>% filter(Survived == 1) %>%
  .$Sex %>%
  table() 
p.female.survived <- p.female.survived[[1]]/sum(p.female.survived)
p.female.survived
```

```{r prob of survived given female}
p.survived.female <- (p.female.survived * p_survived)/p_female
p.survived.female
```

There is a 70.8% of a female surviving based the Bayesian probability. 

```{r}
p.survived.male <- ((1 - p.female.survived) * p_survived) / (1 - p_female)
p.survived.male
```

```{r table of survived by sex}
survived_by_sex <- data.frame(
  sex = c("Female", "Male"),
  prob_survived = c(p.survived.female, p.survived.male),
  prob_sex = c(p_female, 1-p_female)
)
survived_by_sex
```


```{r plot of survived by sex, echo=FALSE}
survived_by_sex %>% ggplot(aes(prob_sex, prob_survived)) +
  geom_label(aes(label = sex, fill = sex)) +
  ylab("Likelihood of Suviving given Sex") +
  xlab("Likelihood of each Sex") +
  guides(fill = F)
```


## Determining probabilities per class
$$ P\left(Survived\ |\ Class_{i}\right) = 
\frac{P\left(Survived\ |\ Class_{i}\right) * P\left(Class_{i}\right)}
{P\left(Survived\right)}$$
### Finding the probability of each class given survived: 

```{r defining a class table given survived}
class.survived <- titanic.clean %>%
  filter(Survived == 1) %>%
  pull(Pclass) %>%
  table()
class.survived
```
```{r class table not dependant on survived}
class <- titanic.clean %>% 
  filter(!is.na(Survived)) %>%
  pull(Pclass) %>%
  table()
class
```
```{r probablitites of each class}
p.class.first <- class[[1]]/sum(class)
p.class.second <- class[[2]]/sum(class)
p.class.third <- class[[3]]/sum(class)
```



```{r probility of first class given survived}
p.first.survived <- class.survived[[1]] / sum(class.survived)
p.first.survived
```
The probability of being in first class given that a passenger survived is 39.76%

```{r probability of second class given survived}
p.second.survived <- class.survived[[2]] / sum(class.survived)
p.second.survived
```
The probability of being in second class given that a passenger survived is 25.43%

```{r probability of third class given survived}
p.third.survived <- class.survived[[3]] / sum(class.survived)
p.third.survived
```
The probability of being in third class given that a passenger survived is 34.79%

### Finding the probablity of surviving in each class:

```{r probability of survived given first class}
p.survived.first <- (p.first.survived * p_survived)/p.class.first
p.survived.first
```

```{r probability of survived given second class}
p.survived.second <- (p.second.survived * p_survived)/p.class.second
p.survived.second
```

```{r probability of survived given thrid class}
p.survived.third <- (p.third.survived * p_survived)/p.class.third
p.survived.third
```

```{r table of survival probabilities given class}
survived_by_class <- data.frame(Class = c("first", "second", "third"),
           survival_prob = c(p.survived.first, p.survived.second, p.survived.third),
           class_prob = c(p.class.first, p.class.second, p.class.third))
survived_by_class
```
```{r plot of survival and class probabilites, echo=FALSE}
survived_by_class %>% ggplot(aes(class_prob, survival_prob)) +
  geom_label(aes(label = Class, fill = Class)) +
  xlab("Likelihood of Being in each Class") +
  ylab("Likelihood of Sruviving in each Class") +
  guides(fill = FALSE)
```

### Finding probabilites for each age group
$$
P\left(Survived\ |\ AgeGroup_i\right) = 
\frac{
P\left(AgeGroup_i\ |\ Survived\right) * P\left(Survived\right)
}{
P\left(AgeGroup_i\right)
}
$$
####Defining the age groups
```{r summary of the Age vector}
summary(titanic.clean$Age)
```

building a function to determine age group. 
```{r defining age group function}
age_group <- function(x){
  if(x< 18){
    group <- "Child"
  } else if (between(x, 18, 35)){
    group <- "Young_Adult"
  } else if (between(x, 35, 65)){
    group <- "Adult"
  } else if (x > 65){
    group <- "Senior"
  }
  return(group)
}
```

Adding the age group vector
```{r adding age group to the table}
titanic.clean <- titanic.clean %>% 
  mutate(Age_Group = sapply(Age, age_group))
head(titanic.clean$Age_Group, n=10)
```
Determining the probability that a passenger would be in each age group.
```{r age group table}
age.table <- titanic.clean %>% 
  filter(!is.na(Survived))%>%
  pull(Age_Group) %>%
  table()
age.table
```
```{r calculating probabilities of each age gropu}
p.age.child <- age.table[[2]]/sum(age.table)
p.age.Yadult <- age.table[[4]]/sum(age.table)
p.age.adult <- age.table[[1]]/sum(age.table)
p.age.senior <- age.table[[3]]/sum(age.table)
```

```{r building table of survived age ranges}
age.survived.table <- titanic.clean %>%
  filter(Survived == 1) %>%
  pull(Age_Group) %>%
  table()
age.survived.table
```

```{r defining probability of age group given survived}
p.child.survived <- age.survived.table[[2]]/sum(age.survived.table)
p.Yadult.survived <- age.survived.table[[4]]/sum(age.survived.table)
p.adult.survived <- age.survived.table[[1]]/sum(age.survived.table)
p.senior.survived <- age.survived.table[[3]]/sum(age.survived.table)
```

```{r probability of survived given age group}
p.survived.child <- (p.child.survived * p_survived) / p.age.child
p.survived.Yadult <- (p.Yadult.survived * p_survived) / p.age.Yadult
p.survived.adult <- (p.adult.survived * p_survived) / p.age.adult
p.survived.senior <- (p.senior.survived * p_survived) / p.age.senior
```

```{r table of survived by age group}
survived_by_age <- data.frame(
  Age_Group = c("Child", "Young Adult", "Adult", "Senior"),
  Survived_Prob = c(p.survived.child, p.survived.Yadult, p.survived.adult, p.survived.senior),
  Age_Prob = c(p.age.child, p.age.Yadult, p.age.adult, p.age.senior)
)
survived_by_age
```

```{r plot of survival given age group}
survived_by_age %>% ggplot(aes(Age_Prob, Survived_Prob))+
  geom_label(aes(label = Age_Group, fill = Age_Group)) + 
  xlab("Likelihood of Age Group") +
  ylab("Likelihood of Survived given Age Group") +
  guides(fill = F)
```


# Testing The Model So Far
building a function to compute survival based on sex and class
```{r compute survival from sex and class}
class_prob <- function(x){
  if(x == 3){
    prob <- p.survived.third
  } else if(x == 2){
    prob <- p.survived.second
  } else if(x == 1){
    prob <- p.survived.first
  }
  return(prob)
}

sex_prob <- function(x){
  if(x == "female"){
    prob <- p.survived.female
  } else if(x == "male"){
    prob <- p.survived.male
  }
  return(prob)
}

age_prob <- function(x){
  if(x == "Child"){
    prob <- p.survived.child
  } else if(x == "Young_Adult"){
    prob <- p.survived.Yadult
  } else if(x == "Adult"){
    prob <- p.survived.adult
  } else if(x == "Senior"){
    prob <- p.survived.senior
  } else {
    prob <- NA
  }
  return(prob)
}
```

```{r reviewing the outcome so far}
titanic.clean <- titanic.clean %>% mutate(
  survived_class_prob = sapply(Pclass, class_prob),
  survived_sex_prob = sapply(Sex, sex_prob),
  survived_age_prob = sapply(Age_Group, age_prob),
  survived_prob = survived_class_prob * survived_sex_prob * survived_age_prob
) %>% 
  select(!c(survived_class_prob, survived_sex_prob, survived_age_prob))
  # filter(survived_prob >.1) %>%
  # filter(!is.na(Survived))%>%
  # pull(Survived)%>%
  # # head(n=10)
  # summary()
  # ggplot(aes(survived_prob, fill = Sex, color = Pclass))+
  # geom_density(aes(y=..density..))
```

If I set the cut off at 10% survival probability, I should be 94% accurate in predicting survival

# Applying the model

creating new data frame with the model applied
```{r building final data frame}
titanic.final <- titanic.clean %>% 
  mutate(Survived = ifelse(survived_prob > .1, 1, 0))
tail(titanic.final, n=20)
```

```{r}
table(titanic.final$Survived)
```

